My title

### 2024-11-23 paper/paper.qmd
update model, data and paper

library(tidyverse)
library(here)
library(arrow)
library(corrplot)
library(knitr)
library(kableExtra)

raw_data_climate <- read_csv(here("data/01-raw_data/climateyvr.csv"))
raw_data_ahccd <- read_csv(here("data/01-raw_data/ahccdyvr.csv"))
analysis_data <- read_parquet(here("data/02-analysis_data/cleaned_data.parquet"))
train_data <- read_parquet(here("data/02-analysis_data/train_data.parquet"))
test_data <- read_parquet(here("data/02-analysis_data/test_data.parquet"))
m1 <- readRDS(here("models/mlr_1.rds"))
m2 <- readRDS(here("models/mlr_2.rds"))
glm_nolog <- readRDS(here("models/glm_nolog.rds"))
glm <- readRDS(here("models/glm_model.rds"))
brm <- readRDS(here("models/brm_model.rds"))
```

@@

# 1. Linearity: Residuals vs Fitted Plot
plot(glm_model, which = 1, main = "Linearity Check: Residuals vs Fitted")
# Look for a random scatter of points. Patterns indicate non-linearity.

# 2. Homoscedasticity: Scale-Location Plot
plot(glm_model, which = 3, main = "Homoscedasticity Check: Scale-Location")
# Points should be evenly spread. A funnel shape indicates heteroscedasticity.

# 3. Normality of Residuals: Q-Q Plot
plot(glm_model, which = 2, main = "Normality Check: Q-Q Plot")
# Points should lie approximately along the diagonal line.

# 4. Residual Histogram: Another Normality Check
residuals <- residuals(glm_model)
hist(residuals, main = "Histogram of Residuals", xlab = "Residuals")
# Check for symmetry. A skewed histogram suggests non-normal residuals.

# 5. Multicollinearity: Variance Inflation Factor (VIF)
library(car)
vif_values <- vif(glm_model)
print(vif_values)
# VIF > 5 indicates high multicollinearity. Consider removing highly correlated predictors.

# 6. Cook's Distance: Influence of Observations
cooksd <- cooks.distance(glm_model)
plot(cooksd, main = "Cook's Distance", type = "h")
abline(h = 4 / nrow(model.frame(glm_model)), col = "red", lty = 2)
# Points above the red line are highly influential.

# 7. Leverage: Identify High Leverage Points
leverage <- hatvalues(glm_model)
plot(leverage, main = "Leverage Check", type = "h")
abline(h = 2 * mean(leverage), col = "red", lty = 2)
# Points above the red line have high leverage.

# 8. Correlation Between Predictors
cor_matrix <- cor(model.frame(glm_model)[, -1]) # Exclude response variable
print(round(cor_matrix, 2))
# Check for highly correlated predictors (e.g., correlation > 0.8 or < -0.8).

@@

The cleaned dataset contains a range of weather variables providing detailed monthly observations. The `date` variable represents the observation month, standardized to the first day of each month. **`wind_speed`** (km/h) captures average monthly wind speeds, while **`total_precipitation`** (mm) measures the total monthly precipitation, including rain and snow. **`snow`** (mm) records total snowfall, and **`pressure_station`** (kPa) indicates atmospheric pressure at the observation station. **`max_temp`** (°C), **`min_temp`** (°C), and **`mean_temp`** (°C) represent the monthly averages of maximum, minimum, and overall temperatures, respectively. **`total_rain`** (mm) focuses solely on rainfall amounts, distinct from snowfall. **`gust_speed_km_h`** (km/h) records the monthly average of maximum gust speeds. Additionally, constructed variables include **`mean_temp_F`**, the mean temperature was converted to Fahrenheit using 
$$
(\text{mean\_temp} \times 1.8) + 32
$$
, and **`log_mean_temp`**, the log-transformed Fahrenheit temperature, was calculated as 
$$
\log(\text{mean\_temp\_F})
$$
. 
Additionally, a Box-Cox transformation was applied to the `total_precipitation` variable to address skewness and stabilize variance, resulting in the new variable `total_precipitation_boxcox`. For `gust_speed_km_h`, a log transformation was used to stabilize variance and reduce right-skewness in its distribution, creating the new variable `log_gust_speed`.

All column names were cleaned and standardized using `janitor` in tidyverse [@tidyverse] to ensure consistency and readability. Dates were parsed into a unified format (`yyyy-mm-dd`) and aligned with monthly observations using the lubridate package [@lubridate]. The datasets were merged into a single combined dataset using the `date_time` variable as the common key. Finally, constructed variables, including `mean_temp_F`, `total_precipitation_boxcox`, `log_gust_speed` and `log_mean_temp`, were added to the cleaned data.

@@

```{r}
#| label: tbl-cleaned-data
#| tbl-cap: Sample of Cleaned Weather Data
#| echo: false
#| warning: false
analysis_data |>
  head(6) |>
  select(
    wind_speed, total_precipitation, snow, pressure_station, 
    max_temp, min_temp, mean_temp, total_rain
  ) |>
  mutate(across(everything(), ~ round(., 2))) |>
  kable(
    col.names = c("Wind Speed", "Total Precip.", "Snow", "Pressure", 
                  "Max Temp", "Min Temp", "Mean Temp", "Rain"),
    align = "c",  # Center-align all columns
    format = "latex"  # Ensure LaTeX output
  )
analysis_data |>
  head(6) |>
  select(
    gust_speed_km_h, log_mean_temp, total_precipitation_boxcox, log_gust_speed
  ) |>
  mutate(across(everything(), ~ round(., 2))) |>
  kable(
    col.names = c("Gust Speed", "Log of Mean Temp", "Box-Cox Total Precip.", "Log of Gust Speed"),
    align = "c",  # Center-align all columns
    format = "latex"  # Ensure LaTeX output
  )
```

@@

## Characteristics of Cleaned Data
### Skewness in Mean Temperature, Total Precipitation and Gust Speed

@@

In @fig-hist-all, Transformations are also applied to Total Precipitation and Gust Speed. Total Precipitation has a strong right skew, with most values low and a few extreme high values. A logarithmic transformation is used to compress the large values, reducing skewness and making the distribution more symmetric. For Gust Speed, a Box-Cox transformation is applied to adjust its moderate skewness. This transformation reshaped the data to better approximate a normal distribution. These adjustments improve the suitability of these variables for statistical analyses that assume normality.

@@

```{r}
#| label: fig-boxplot-snow
#| fig-cap: Total Snow shows Zero Inflation
#| echo: false
#| warning: false
boxplot(analysis_data$snow,
        horizontal = TRUE,
        main = "Boxplot of Total Snow",
        ylab = "Total Snow",
        xlab = "Snow (in cm)", 
        col = "lightblue",
        notch = TRUE)
```

### Some Variables Has Strong Linear Relationships

@@

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to investigate... Background details and diagnostics are included in [Appendix -@sec-model-details].

@@

```{r}
fitted_values <- fitted(m2)
ggplot(data = train_data, aes(x = fitted_values, y = mean_temp_F)) +
  geom_point(color = "blue", alpha = 0.7) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed", size = 1) +
  labs(x = "Predicted Temperature", y = "Actual Temperature", title = "Actual Temperature vs. Predicted Values")
```

@@

# Summary Statistic of cleaned dataset
\newpage
```{r}
#| label: tbl-summary
#| tbl-cap: Summary Statistics of Raw Climate Data 
#| echo: false
#| warning: false
# Generate the table
analysis_data |>
  select(wind_speed, total_precipitation, snow, pressure_station,
  max_temp, min_temp) |>
  summary() |>
  kable(
    col.names = c(
      "Wind Speed", "Total Precipitation", "Snow", "Pres.", 
      "Max Temp", "Min Temp"
    ),
    booktabs = TRUE,
    align = "c",  # Center-align all columns
    format = "latex"  # Ensure LaTeX output
  )
analysis_data |>
  select(mean_temp, total_rain, gust_speed_km_h,
  mean_temp_F, log_mean_temp) |>
  summary() |>
  kable(
    col.names = c(
      "Mean Temp", "Rain", 
      "Max Gust Speed", "Mean Temp in F", "Log of Mean Temp"
    ),
    booktabs = TRUE,
    align = "c",  # Center-align all columns
    format = "latex"  # Ensure LaTeX output
  )
analysis_data |>
  select(total_precipitation_boxcox, log_gust_speed) |>
  summary() |>
  kable(
    col.names = c(
      "Box Cox of Precipitation", "Log of Gust Speed"
    ),
    booktabs = TRUE,
    align = "c",  # Center-align all columns
    format = "latex"  # Ensure LaTeX output
  )
```
\newpage

```{r}
#| label: fig-hist-all
#| fig-cap: Other Variables Show Normal Distribution
#| fig-subcap: ['Wind Speed Shows Symmetric Distribution', 'Total Precipitation Shows Normal Distribution but Skewed Towards Low Values', 'Pressure Station Has a Nearly Normal Distribution', 'Gust Speed Shows Normal Distribution and Moderate Skewness', 'Box-Cox Transformed Gust Speed shows Reduces Skewness', 'Log Transformed Gust Speed shows Reduces Skewness']
#| echo: false
#| warning: false
#| layout-ncol: 2

@@

@@

```{r}
# Fit a model

>>>>

library(tidyverse)
library(here)
library(arrow)
library(corrplot)
library(knitr)
library(kableExtra)
library(car)

raw_data_climate <- read_csv(here("data/01-raw_data/climateyvr.csv"))
raw_data_ahccd <- read_csv(here("data/01-raw_data/ahccdyvr.csv"))
analysis_data <- read_parquet(here("data/02-analysis_data/cleaned_data.parquet"))
train_data <- read_parquet(here("data/02-analysis_data/train_data.parquet"))
test_data <- read_parquet(here("data/02-analysis_data/test_data.parquet"))
m1 <- readRDS(here("models/m1.rds"))
m4 <- readRDS(here("models/m4.rds"))
glm <- readRDS(here("models/glm_poly.rds"))
brm <- readRDS(here("models/brm.rds"))
glm_log <- readRDS(here("models/glm_log.rds"))
```

@@

# 1. Linearity: Residuals vs Fitted Plot
plot(glm_model, which = 1, main = "Linearity Check: Residuals vs Fitted")
plot(glm_log, which = 1, main = "Linearity Check: Residuals vs Fitted")

# Look for a random scatter of points. Patterns indicate non-linearity.

# 2. Homoscedasticity: Scale-Location Plot
plot(glm_model, which = 3, main = "Homoscedasticity Check: Scale-Location")
plot(glm_log, which = 3, main = "Homoscedasticity Check: Scale-Location")
# Points should be evenly spread. A funnel shape indicates heteroscedasticity.

# 3. Normality of Residuals: Q-Q Plot
plot(glm_model, which = 2, main = "Normality Check: Q-Q Plot")
plot(glm_log, which = 2, main = "Normality Check: Q-Q Plot")

# Points should lie approximately along the diagonal line.

# 4. Residual Histogram: Another Normality Check
residuals <- residuals(glm_model)
hist(residuals, main = "Histogram of Residuals", xlab = "Residuals")
residuals <- residuals(glm_log)
hist(residuals, main = "Histogram of Residuals", xlab = "Residuals")

# Check for symmetry. A skewed histogram suggests non-normal residuals.

# 5. Multicollinearity: Variance Inflation Factor (VIF)
vif_values <- vif(glm_model)
print(vif_values)
vif_values <- vif(glm_log)
print(vif_values)
# VIF > 5 indicates high multicollinearity. Consider removing highly correlated predictors.

# 6. Cook's Distance: Influence of Observations
cooksd <- cooks.distance(glm_model)
plot(cooksd, main = "Cook's Distance", type = "h")
abline(h = 4 / nrow(model.frame(glm_model)), col = "red", lty = 2)

cooksd <- cooks.distance(glm_log)
plot(cooksd, main = "Cook's Distance", type = "h")
abline(h = 4 / nrow(model.frame(glm_log)), col = "red", lty = 2)
# Points above the red line are highly influential.

# 7. Leverage: Identify High Leverage Points
leverage <- hatvalues(glm_model)
plot(leverage, main = "Leverage Check", type = "h")
abline(h = 2 * mean(leverage), col = "red", lty = 2)

leverage <- hatvalues(glm_log)
plot(leverage, main = "Leverage Check", type = "h")
abline(h = 2 * mean(leverage), col = "red", lty = 2)
# Points above the red line have high leverage.

@@

The cleaned dataset contains a range of weather variables providing detailed monthly observations. The `date` variable represents the observation month, standardized to the first day of each month. **`wind_speed`** (km/h) captures average monthly wind speeds, while **`total_precipitation`** (mm) measures the total monthly precipitation, including rain and snow. **`snow`** (mm) records total snowfall, and **`pressure_station`** (kPa) indicates atmospheric pressure at the observation station. **`max_temp`** (°C), **`min_temp`** (°C), and **`mean_temp`** (°C) represent the monthly averages of maximum, minimum, and overall temperatures, respectively. **`total_rain`** (mm) focuses solely on rainfall amounts, distinct from snowfall. **`gust_speed_km_h`** (km/h) records the monthly average of maximum gust speeds. Additionally, constructed variables include **`mean_temp_F`**, the mean temperature was converted to Fahrenheit using 
$$
(\text{mean\_temp} \times 1.8) + 32
$$
, and **`log_mean_temp`**, the log-transformed Fahrenheit temperature, was calculated as 
$$
\log(\text{mean\_temp\_F})
$$
. 
Additionally, a Box-Cox transformation was applied to the `total_precipitation` variable to address skewness and stabilize variance, resulting in the new variable `total_precipitation_boxcox`. For `gust_speed_km_h`, `wind_speed` and `pressure_station`, a log transformation was used to stabilize variance and reduce right-skewness in their distribution, creating the new variable `log_gust_speed`, `log_wind_speed` and `log_pressure`.

All column names were cleaned and standardized using `janitor` in tidyverse [@tidyverse] to ensure consistency and readability. Dates were parsed into a unified format (`yyyy-mm-dd`) and aligned with monthly observations using the lubridate package [@lubridate]. The datasets were merged into a single combined dataset using the `date_time` variable as the common key. Finally, constructed variables, including `mean_temp_F`, `total_precipitation_boxcox`, `log_gust_speed`, `log_mean_temp`, `log_wind_speed` and `log_pressure` were added to the cleaned data.

@@

```{r}
#| label: tbl-cleaned-data
#| tbl-cap: Sample of Cleaned Weather Data
#| tbl-subcap: ['Sample of Cleaned Weather Data', 'Gust Speed and Transformed Data']
#| echo: false
#| warning: false
analysis_data |>
  head(6) |>
  select(
    wind_speed, total_precipitation, snow, pressure_station, 
    max_temp, min_temp, mean_temp, total_rain
  ) |>
  mutate(across(everything(), ~ round(., 2))) |>
  kable(
    col.names = c("Wind Speed", "Total Precip.", "Snow", "Pressure", 
                  "Max Temp", "Min Temp", "Mean Temp", "Rain"),
    format = "latex", booktabs = TRUE,
  )

analysis_data |>
  head(6) |>
  select(log_gust_speed, log_wind_speed, log_pressure
  ) |>
  mutate(across(everything(), ~ round(., 2))) |>
  kable(
    col.names = c("Log of Gust Speed", "Log of Wind Speed", "Log of Pressire"),
    format = "latex", booktabs = TRUE,
  )
analysis_data |>
  head(6) |>
  select(
    gust_speed_km_h, log_mean_temp, total_precipitation_boxcox
  ) |>
  mutate(across(everything(), ~ round(., 2))) |>
  kable(
    col.names = c("Gust Speed", "Log of Mean Temp", "Box-Cox Total Precip."),
    format = "latex", booktabs = TRUE,
  )
```

@@

## Characteristics of Cleaned Data

All of the variables in the dataset are numeric, the histograms are plotted in @fig-hist-temp and @fig-hist-all. The following section explains the characteristics of these variables.

### Skewness in Mean Temperature, Total Precipitation and Gust Speed

@@

In @fig-hist-all, Transformations are also applied to Wind Speed, Pressure, Total Precipitation and Gust Speed. Total Precipitation has a strong right skew, with most values low and a few extreme high values. We apply log transformation to predictors including wind speed, pressure and precipitation. For Gust Speed, a Box-Cox transformation is applied to adjust its moderate skewness. This transformation reshaped the data to better approximate a normal distribution. These adjustments improve the suitability of these variables for statistical analyses that assume normality.

@@

```{r}
#| label: fig-boxplot-snow
#| fig-cap: Total Snow shows Zero Inflation
#| fig-subcap: ['A Large Proportion of Observations With No Snowfall', 'Prevalence of Zero Values and a Skewed Pattern in the Non-zero Snowfall Measurements']
#| echo: false
#| warning: false
#| layout-ncol: 2
boxplot(analysis_data$snow,
        horizontal = TRUE,
        main = "Boxplot of Total Snow",
        ylab = "Total Snow",
        xlab = "Snow (in cm)", 
        col = "lightblue",
        notch = TRUE)

ggplot(analysis_data, aes(x = snow)) +
  geom_histogram(fill = "red", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Snow", x = "Snow (cm)", y = "Frequency") +
  theme_minimal()

### Variables with Strong Linear Relationships

@@

The goal of our modelling strategy is to find a model that can predict temperature changes with other weather data.

We build a bayesian model, a general linear model and a general linear model with a 2 degrees of polynomial transformation. We determine the best model is the linear model with polynomial transformation. The detailed steps are recorded in @sec-model-select.

@@

```{r}
fitted_values <- fitted(m4)
ggplot(data = train_data, aes(x = fitted_values, y = mean_temp_F)) +
  geom_point(color = "blue", alpha = 0.7) +
  geom_abline(intercept = 0, slope = 1, color = "red", linetype = "dashed", size = 1) +
  labs(x = "Predicted Temperature", y = "Actual Temperature", title = "Actual Temperature vs. Predicted Values")
```

@@

# Summary Statistic of cleaned dataset
\newpage
```{r}
#| label: tbl-summary
#| tbl-cap: Summary Statistics of Raw Climate Data 
#| echo: false
#| warning: false
# Generate the table
analysis_data |>
  select(wind_speed, total_precipitation, snow, pressure_station,
  max_temp, min_temp) |>
  summary() |>
  kable(
    col.names = c(
      "Wind Speed", "Total Precipitation", "Snow", "Pres.", 
      "Max Temp", "Min Temp"
    ),
    booktabs = TRUE,
    align = "c",  # Center-align all columns
    format = "latex"  # Ensure LaTeX output
  )
analysis_data |>
  select(mean_temp, total_rain, gust_speed_km_h,
  mean_temp_F, log_mean_temp) |>
  summary() |>
  kable(
    col.names = c(
      "Mean Temp", "Rain", 
      "Max Gust Speed", "Mean Temp in F", "Log of Mean Temp"
    ),
    booktabs = TRUE,
    align = "c",  # Center-align all columns
    format = "latex"  # Ensure LaTeX output
  )
analysis_data |>
  select(total_precipitation_boxcox, log_gust_speed, log_wind_speed, log_pressure) |>
  summary() |>
  kable(
    col.names = c(
      "Box Cox of Precipitation", "Log of Gust Speed", "Log of Wind Speed",
      "Log of Pressure"
    ),
    booktabs = TRUE,
    align = "c",  # Center-align all columns
    format = "latex"  # Ensure LaTeX output
  )
```
\newpage

```{r}
#| label: fig-hist-all
#| fig-cap: Other Variables Show Normal Distribution
#| fig-subcap: ['Wind Speed Shows Symmetric Distribution', 'Total Precipitation Shows Normal Distribution but Skewed Towards Low Values', 'Pressure Station Has a Nearly Normal Distribution', 'Gust Speed Shows Normal Distribution and Moderate Skewness', 'Box-Cox Transformed Precipitation shows Reduces Skewness', 'Log Transformed Gust Speed shows Reduces Skewness', 'Log Transformed Pressure shows Reduces Skewness', 'Log Transformed Wind Speed shows Reduces Skewness']
#| echo: false
#| warning: false
#| layout-ncol: 3

@@

# Plot histogram for log_pressure
ggplot(df, aes(x = log_pressure)) +
  geom_histogram(fill = "black", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Log of Pressure Station", x = "Pressure (hPa)", y = "Frequency") +
  theme_minimal()

# Plot histogram for wind_speed
ggplot(df, aes(x = log_wind_speed)) +
  geom_histogram(fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Log of Wind Speed", x = "Wind Speed (km/h)", y = "Frequency") +
  theme_minimal()

@@

# Additional Model Detail {#sec-model-select}
